Advice on describing Bayesian analysis of neutron and X-ray reflectometry

### 2022-07-21 src/tex/ms.tex
Deal with all of Andrew Nelson's comments

Reflectometry analysis can be described, in the most simplistic terms, as a comparison and refinement of a model based on some parameters, $\mathbf{x}$, to reproduce some reflectivity data set, $\mathbf{D}$. 
This refinement process involves comparing the model to the data and calculating some goodness-of-fit value or likelihood, $p(\mathbf{D} | \mathbf{x})$, and modifying the model to optimize the goodness-of-fit or maximise the likelihood.
A commonly used goodness-of-fit parameter is the $\chi^2$ parameter which is found as \cite{nelson_refnx_2019}, 
%
\begin{equation}
    \chi^2 = \sum_{q=q_{\text{min}}}^{q_{\text{max}}}{\bigg[\frac{(R(q) - R(q)_{\text{m}})}{\sigma_R(q)}\bigg]^2}, 
\end{equation}
%
where, $R(q)$ and $R(q)_{\text{m}}$ are the measured and modelled reflectivity at a given $q$, the measured momentum transfer vector, and $\sigma_R(q)$ is the uncertainty associated with the measured reflectivity at each $q$ point. 
Under an assumption of normally distributed residuals $R(q) - R(q)_{\text{m}} \sim \mathcal{N}(0, \sigma_R(q))$, the likelihood is related to the $\chi^2$ variable in the following way:
%
\begin{equation}
    \ln[p(\mathbf{D} | \mathbf{x})] = -\frac{1}{2} \bigg(\chi^2 + \sum_{q=q_{\text{min}}}^{q_{\text{max}}}\ln{\big[2\pi\sigma_R(q)^2\big]}\bigg).
    \label{equ:likelihood}
\end{equation}
%
The input for this refinement process is the model and some initial parameter values, which may be an absolute value or some parameter range, depending on the refinement algorithm. 
% ARJN the output of a Bayesian modelling process is a probability distribution for x. If everything is nice these will be Normally distributed, but not necessarily. We then report some statistical measures, e.g. mode/mean/median.
The output is a set of values for $\mathbf{x}$, potentially with associated error bars, where these are present they typically describe a standard deviation from the mean of a Gaussian probability distribution. 
% ARJN Perhaps a Bayesian modelling process can deal with inaccurate reporting of uncertainties, e.g. some factor that accounts for error bars that are statistically underestimated.
This process implicitly assumes that the data is completely reduced, accounting for all experimental parameters, uncertainties are accurately described and the model can accurately describe the data.

@@

The optimised parameters from the minimisation algorithm, which depend on the particular algorithm, often include some statistical uncertainty. 
% ARJN, parameters uncertainties only assumed to be normally distributed for gradient-based minimizers. Bayesian analyses don't assume any form of statistical distribution.
% Provide a reference for how statistical uncertainties are found, e.g. the appropriate place in Bevington.
How these statistical uncertainties are found is beyond the scope of this work, but it is important to acknowledge that this uncertainty typically assumes that the probability distribution of the parameter is Gaussian in nature. 
This probability distribution is either the partial likelihood or posterior, the latter when some prior is included and Bayes theorem is applied. The posterior describes our understanding of the parameter values given the data that was measured.
When Bayesian modelling is used and the prior is included, the posterior probability is found as, 
%
\begin{equation}
  p(\mathbf{x} | \mathbf{D}) \propto p(\mathbf{D} | \mathbf{x}) p(\mathbf{x}).
  \label{equ:bayes}
\end{equation}
%
Therefore, when Bayesian modelling is performed, the priors and likelihood are of fundamental importance to the results that are obtained (the posterior), and any scientific conclusions that are drawn.

@@

When a prior probability can be described with a mathematical function, this should be done by providing this function in the clearest possible language. 
For example, if the prior is taken from a single complementary measurement that is defined as a value with some uncertainty, which represents a normal distribution with a mean and standard deviation, this information should be provided. 
This is shown in Figure~\ref{fig:prior} for the density of silicon nitride (\ce{Si3N4}) produced by atomic layer deposition \cite{knoops_atomic_2015}) which is used to inform the value for a scattering length density for some layer of the material.
% ARJN I would argue that reporting a prior graphically is not really all that useful, unless it's obtained by sampling.
% ARJN I reckon the concise version is what we should recommend. Reporting the equation for the prior is overkill for a normal distribution, and may only be necessary where there's vagueness about a distribution, e.g. there's a few ways of describing a log-normal.
Such a prior probability distribution could be described graphically, Figure~\ref{fig:prior}, or in prose as being ``normally distributed with a mean of \SI{2.9}{\gram\per\cubic\centi\meter} and a standard deviation of \SI{0.1}{\gram\per\cubic\centi\meter}'', mathematically as, 
%
\begin{equation}
  p(\rho_m) = \frac{1}{\sigma\sqrt{2\pi}}\exp\Bigg[-\frac{1}{2}\bigg(\frac{\rho_m-\mu}{\sigma}\bigg)^2\Bigg]
\end{equation}
%
where, $\mu=\SI{2.9}{\gram\per\cubic\centi\meter}$ and $\sigma=\SI{0.1}{\gram\per\cubic\centi\meter}$, or more concisely $p(\rho_m) \sim \mathcal{N}(\mu=\SI{2.9}{\gram\per\cubic\centi\meter}, \sigma=\SI{0.1}{\gram\per\cubic\centi\meter})$. 
The same descriptive approach could be taken for any common statistical distribution, including log-normal or truncated normal distributions. 
%
\begin{figure}
  \script{prior.py}
  \includegraphics[width=\columnwidth]{figures/prior.pdf}
  \caption{
    A prior probability distribution for \ce{Si3N4} with a density of $\rho_m=\SI{2.9\pm0.1}{\gram\per\cubic\centi\meter}$. 
  }
  \label{fig:prior}
\end{figure}
%

It is possible that the prior distribution cannot be described with a simple mathematical function, with multi-modal priors being an example.
%, if it is a multimodal model result from some other sampling method, then the chain from the prior sampling should be given. 
%The chain is all of the samples investigated in the sampling and should be shared, although in the case that this chain is very large a subsampled object may be shared, in which case, the autocorrelation analysis performed should be described (see Section~\ref{sec:posterior} for a more complete discussion of this).
To use such a prior probability in Bayesian analysis, some functional description of the prior must be generated, most commonly this will be some kernel density estimation, when this is used it is also necessary to state the structure of the kernel being used. 
An example of this is shown in Figure~\ref{fig:samples}, where the prior probability for the volume of a phospholipid tail could be found from molecular dynamics simulation where there are three common conformers that the lipid is likely to have.
%
\begin{figure}
  \script{samples.py}
  \includegraphics[width=\columnwidth]{figures/samples.pdf}
  \caption{
    A hypothetical prior probability distribution for a DPPC lipid that could arise from a molecular dynamics simulation (orange histogram) and a Gaussian kernel density estimation for the probability distribution using a bandwidth factor of \num{0.05} (blue line). 
  }
  \label{fig:samples}
\end{figure}
%

@@

Bayes theorem (given in Equation~\ref{equ:bayes}) consists of the product of the prior and the likelihood. 
The former describes our current understanding of the parameters before we conduct some experiments, while the latter describes how well the data is described by the model parameters. 
% ARJN we're not offering the reader suggestions on how to state this. Can they just state that they used packageX with version Y with likelihood option Z?
% ARJN there are multiple ways of weighting data, I'm not sure most users would know the details of how the likelihood is calculated in a given package.
% ARJN Should package authors state how this is done more comprehensively?
Therefore, we must state how the likelihood for a given model and data is calculated, in particular, because although Equation~\ref{equ:likelihood} is a common approach, it assumes a normally distributed uncertainty for a measured reflectivity value. 
% ARJN I think this is something like less than 30 (10?)???
While a normally distributed uncertainty is the most common, it may not be accurate in all circumstances, for example, when low numbers of counts are present, a Poisson uncertainty may offer a more accurate description.

@@

Bayesian analysis methods typically involve using some sampling process, such as Markov chain Monte Carlo, to estimate the posterior probability distributions for each of the parameters. 
Assuming there are $m$ parameters under investigation, the posterior will be an $m$-dimensional probability distribution. 
The result of a Bayesian sampling process is a ``chain'' consisting of $n$ samples for each parameter. 
Therefore, the full chain has a shape $(m, n)$. 
% ARJN a reader may not know what an 'independent', or 'non-independent sample' is.
% ARJN provide a reference for autocorrelation analysis.
Typically these are histogrammed to show the probability of different values of the parameters. However, to identify independent (non-correlated) samples in the chain, autocorrelation analysis may be performed and the chain ``thinned''. 
We will not cover in detail autocorrelation analysis other than to say that it helps to identify the length of separation required for samples to be independent and that thinning means that we only included samples separated by this length in the final chain.
Additionally, it is valuable to report the use of convergence diagnostics, such as the Gelman-Rubin statistic \cite{gelman_inference_1992}, which can assist in determining if a chain appropriately describes a posterior.

@@

To report values for parameters and some form of statistical uncertainty, two approaches can be taken from the posterior chain. 
The first is to use some known statistical distribution that describes the samples well.
% ARJN I think this will rapidly become intractible. For example what would one do with e.g. 100 parameters, should one be checking them all for normality? There will inevitably ones that are skewed. Then there will be a hodgepodge of answers.
% ARJN THe only point in checking for normality (or matching any distribution) would be if the science was predicting some specific distribution of parameter values. 
% ARJN For a unimodal distribution I report 50th percentile (median) and half the [16, 84] range.
This is best defined for a normal distribution, for which there are statistical tests to check normality, such as the D'Agostino and Pearson's test \cite{dagostino_omnibus_1971,dagostino_tests_1973} (which is available in the SciPy library as \texttt{scipy.stats.normaltest} \cite{virtanen_scipy_2020}).
As with all statistical tests, this requires some threshold value to be defined to reject the null hypothesis, for this value we recommend \num{0.001} but accept that this is at the discretion of the user. 
If the parameter distribution passes a statistical test for a given distribution type, this can be quoted in the work, with information about the distribution type and the threshold value used, and the distribution can be described based on fitted parameters of the distribution as is discussed above for the Gaussian distribution. 
For example, the three parameters in Figure~\ref{fig:post} pass this statistical tests, with $p$-values of greater than \num{0.01} when \num{1000} random samples are used, therefore we can quote the parameters as normal distributions; \variable{output/mag_sld.txt}\unskip, \variable{output/density.txt}\unskip, and \variable{output/thick.txt}\unskip.

>>>>

Reflectometry analysis can be described, in the most simplistic terms, as a comparison and refinement of a model based on some parameters, $\mathbf{x}$, to reproduce some reflectivity data set, $\mathbf{D}$. 
This refinement process involves comparing the model to the data and calculating some goodness-of-fit value or likelihood, $p(\mathbf{D} | \mathbf{x})$, and modifying the model to optimize the goodness-of-fit or maximise the likelihood.
A commonly used goodness-of-fit parameter is the $\chi^2$ parameter which is found as \cite{nelson_refnx_2019}, 
%
\begin{equation}
    \chi^2 = \sum_{q=q_{\text{min}}}^{q_{\text{max}}}{\bigg[\frac{(R(q) - R(q)_{\text{m}})}{\sigma_R(q)}\bigg]^2}, 
\end{equation}
%
where, $R(q)$ and $R(q)_{\text{m}}$ are the measured and modelled reflectivity at a given $q$, the measured momentum transfer vector, and $\sigma_R(q)$ is the uncertainty associated with the measured reflectivity at each $q$ point. 
Under an assumption of normally distributed residuals $R(q) - R(q)_{\text{m}} \sim \mathcal{N}(0, \sigma_R(q))$, the likelihood is related to the $\chi^2$ variable in the following way:
%
\begin{equation}
    \ln[p(\mathbf{D} | \mathbf{x})] = -\frac{1}{2} \bigg(\chi^2 + \sum_{q=q_{\text{min}}}^{q_{\text{max}}}\ln{\big[2\pi\sigma_R(q)^2\big]}\bigg).
    \label{equ:likelihood}
\end{equation}
%
The input for this refinement process is the model and some initial parameter values, which may be an absolute value or some parameter range, depending on the refinement algorithm.
While for some Bayesian sampling process, the input is a probability distribution for each parameter, known as the prior.
The output is a set of values for $\mathbf{x}$, potentially with associated error bars, where these are present they typically describe a standard deviation from the mean of a Gaussian probability distribution. 
A Bayesian sampling process gives a probability distribution, the posterior, the defines the relative likelihood of different values of each parameter, from this we can report statistical measures, e.g. mode/mean/median.
This process implicitly assumes that the data is completely reduced, accounting for all experimental parameters, uncertainties are accurately described and the model can accurately describe the data.

@@

The optimised parameters from the minimisation algorithm, which depend on the particular algorithm, often include some statistical uncertainty.
This uncertainty comes from an assumption of normally distributed parameters \cite{bevington_data_2002}, however, Bayesian sampling approaches make no assumption of an underlying statistical distribution.
How these statistical uncertainties are found is beyond the scope of this work, but it is important to acknowledge that this uncertainty typically assumes that the probability distribution of the parameter is Gaussian in nature. 
This probability distribution is either the partial likelihood or posterior, the latter when some prior is included and Bayes theorem is applied. The posterior describes our understanding of the parameter values given the data that was measured.
When Bayesian modelling is used and the prior is included, the posterior probability is found as, 
%
\begin{equation}
  p(\mathbf{x} | \mathbf{D}) \propto p(\mathbf{D} | \mathbf{x}) p(\mathbf{x}).
  \label{equ:bayes}
\end{equation}
%
Therefore, when Bayesian modelling is performed, the priors and likelihood are of fundamental importance to the results that are obtained (the posterior), and any scientific conclusions that are drawn.

@@

When a prior probability can be described with a mathematical function, this should be done by providing this function in the clearest possible language. 
For example, if the prior is taken from a single complementary measurement that is defined as a value with some uncertainty, which represents a normal distribution with a mean and standard deviation, this information should be provided. 
This is shown in Figure~\ref{fig:prior} for the density of silicon nitride (\ce{Si3N4}) produced by atomic layer deposition \cite{knoops_atomic_2015}) which is used to inform the value for a scattering length density for some layer of the material.
Such a prior probability distribution could be described graphically, Figure~\ref{fig:prior}, or in prose as being ``normally distributed with a mean of \SI{2.9}{\gram\per\cubic\centi\meter} and a standard deviation of \SI{0.1}{\gram\per\cubic\centi\meter}'', mathematically as, 
%
\begin{equation}
  p(\rho_m) = \frac{1}{\sigma\sqrt{2\pi}}\exp\Bigg[-\frac{1}{2}\bigg(\frac{\rho_m-\mu}{\sigma}\bigg)^2\Bigg]
\end{equation}
%
where, $\mu=\SI{2.9}{\gram\per\cubic\centi\meter}$ and $\sigma=\SI{0.1}{\gram\per\cubic\centi\meter}$, or more concisely $p(\rho_m) \sim \mathcal{N}(\mu=\SI{2.9}{\gram\per\cubic\centi\meter}, \sigma=\SI{0.1}{\gram\per\cubic\centi\meter})$. 
The same descriptive approach could be taken for any common statistical distribution, including log-normal or truncated normal distributions. 
%
\begin{figure}
  \script{prior.py}
  \includegraphics[width=\columnwidth]{figures/prior.pdf}
  \caption{
    A prior probability distribution for \ce{Si3N4} with a density of $\rho_m=\SI{2.9\pm0.1}{\gram\per\cubic\centi\meter}$. 
  }
  \label{fig:prior}
\end{figure}
%

It is possible that the prior distribution cannot be described with a simple mathematical function, with multi-modal priors being an example, if it is a multimodal model result from some other sampling method, then the chain from the prior sampling should be given. 
The chain is all of the samples investigated in the sampling and should be shared, although in the case that this chain is very large a subsampled object may be shared, in which case, the autocorrelation analysis performed should be described (see Section~\ref{sec:posterior} for a more complete discussion of this).
To use such a prior probability in Bayesian analysis, some functional description of the prior must be generated, most commonly this will be some kernel density estimation, when this is used it is also necessary to state the structure of the kernel being used. 
An example of this is shown in Figure~\ref{fig:samples}, where the prior probability for the volume of a phospholipid tail could be found from molecular dynamics simulation where there are three common conformers that the lipid is likely to have.
%
\begin{figure}
  \script{samples.py}
  \includegraphics[width=\columnwidth]{figures/samples.pdf}
  \caption{
    A hypothetical prior probability distribution for a DPPC lipid that could arise from a molecular dynamics simulation (orange histogram) and a Gaussian kernel density estimation for the probability distribution using a bandwidth factor of \num{0.05} (blue line). 
  }
  \label{fig:samples}
\end{figure}
%

@@

Bayes theorem (given in Equation~\ref{equ:bayes}) consists of the product of the prior and the likelihood. 
The former describes our current understanding of the parameters before we conduct some experiments, while the latter describes how well the data is described by the model parameters. 
Therefore, we must state how the likelihood for a given model and data is calculated (giving the analysis package used, version number, and if not the default the likelihood option), in particular, because although Equation~\ref{equ:likelihood} is a common approach, it assumes a normally distributed uncertainty for a measured reflectivity value. 
While a normally distributed uncertainty is the most common, it may not be accurate in all circumstances, for example, when low numbers of counts are present, a Poisson uncertainty may offer a more accurate description.

@@

Bayesian analysis methods typically involve using some sampling process, such as Markov chain Monte Carlo, to estimate the posterior probability distributions for each of the parameters. 
Assuming there are $m$ parameters under investigation, the posterior will be an $m$-dimensional probability distribution. 
The result of a Bayesian sampling process is a ``chain'' consisting of $n$ samples for each parameter. 
Therefore, the full chain has a shape $(m, n)$. 
Typically these are histogrammed to show the probability of different values of the parameters. 
However, to identify independent (non-correlated) samples in the chain, autocorrelation analysis \cite{sokal_monte_1997} may be performed and the chain ``thinned''. 
We will not cover in detail autocorrelation analysis other than to say that it helps to identify the length of separation required for samples to be independent and that thinning means that we only included samples separated by this length in the final chain.
Additionally, it is valuable to report the use of convergence diagnostics, such as the Gelman-Rubin statistic \cite{gelman_inference_1992}, which can assist in determining if a chain appropriately describes a posterior.

@@

To report values for parameters and some form of statistical uncertainty, two approaches can be taken from the posterior chain. 
The first is to use some known statistical distribution that describes the samples well.
This is best defined for a normal distribution, for which there are statistical tests to check normality, such as the D'Agostino and Pearson's test \cite{dagostino_omnibus_1971,dagostino_tests_1973} (which is available in the SciPy library as \texttt{scipy.stats.normaltest} \cite{virtanen_scipy_2020}).
As with all statistical tests, this requires some threshold value to be defined to reject the null hypothesis, for this value we recommend \num{0.001} but accept that this is at the discretion of the user. 
If the parameter distribution passes a statistical test for a given distribution type, this can be quoted in the work, with information about the distribution type and the threshold value used, and the distribution can be described based on fitted parameters of the distribution as is discussed above for the Gaussian distribution. 
For example, the three parameters in Figure~\ref{fig:post} pass this statistical tests, with $p$-values of greater than \num{0.01} when \num{1000} random samples are used, therefore we can quote the parameters as normal distributions; \variable{output/mag_sld.txt}\unskip, \variable{output/density.txt}\unskip, and \variable{output/thick.txt}\unskip.
